---
title: "Weather Analysis"
output: html_document
---
## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Data
The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

* [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb]

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

## Data Processing
Download the file
```{r}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "StormData.csv.bz2", method = "curl")
```

Read the data from the file
```{r}
stormData <- read.csv("StormData.csv.bz2")
```

Load the required libraries
```{r, message = FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)
```

## Calculations
The first calculation will be how many injuries occur by each event.
```{r}
injuries_by_event <- aggregate(INJURIES ~ EVTYPE, data = stormData, sum)
injuries_by_event <- injuries_by_event[order(injuries_by_event$INJURIES, decreasing = TRUE),]
```

Here are the top 10 events that cause injuries.
```{r}
top_injuries <- injuries_by_event[1:10, ]
top_injuries
```

The next calculation is how many deaths occur by each event.
```{r}
deaths_by_event <- aggregate(FATALITIES ~ EVTYPE, data = stormData, sum) 
deaths_by_event <- deaths_by_event[order(deaths_by_event$FATALITIES, decreasing = TRUE), ]
```

Here are the top 10 events that cause deaths.
```{r}
top_deaths <- deaths_by_event[1:10, ]
top_deaths
```

The last calculation is how much damage is done to property and crops by each event.
```{r}
economic_by_event <- aggregate(PROPDMG + CROPDMG ~ EVTYPE, data = stormData, sum)
economic_by_event <- economic_by_event[order(economic_by_event$`PROPDMG + CROPDMG`, decreasing = TRUE), ]
```

Here are the top 15 events that cause the most damage to property and crops.
```{r}
top_economic <- economic_by_event[1:15, ]
top_economic
```

## Results
1. Across the United States, which types of events are most harmful with respect to population health?

These two bargraphs show which events have caused the most injuries and the most deaths to human. Clearly, tornadoes cause the most injuries and most deaths to humans. 
```{r}
plot_injuries <- ggplot(top_injuries, aes(EVTYPE, INJURIES)) + 
        geom_bar(stat = "identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
        ggtitle("Injuries by Event") +
        xlab("Event")

plot_deaths <- ggplot(top_deaths, aes(EVTYPE, FATALITIES)) +
        geom_bar(stat = "identity") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
        ggtitle("Deaths by Event") +
        xlab("Event")

grid.arrange(plot_injuries, plot_deaths, nrow = 1)
```

2. Across the United States, which types of events have the greatest economic consequences?

This bargraph shows how much economic damage each event has on property and crops. Just like the previous question, tornadoes cause the most economic damage.
```{r}
ggplot(top_economic, aes(EVTYPE, `PROPDMG + CROPDMG`)) +
        geom_bar(stat = "identity") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
        ggtitle("Economic Damage") +
        xlab("Event") + 
        ylab("Total Damage")
```



